---
title: "0_Fit_Bit_Tracker_Data_Sorting_and_Filtering"
author: "Jeronimo Miranda"
date: '2023-05-05'
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(skimr)
```

```{r directory setup, include = FALSE}
rstudioapi::getActiveDocumentContext
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```

## Loading the data
We are using the data from <https://www.kaggle.com/datasets/arashnic/fitbit>. 
Assume the path to the data is in a folder above where the repository is.

```{r data loading}
data_path <- "../Fitabase Data 4.12.16-5.12.16/"
dailyActivity <- read_csv(paste0(data_path,"dailyActivity_merged.csv"), 
    col_types = cols(Id = col_character(), 
        ActivityDate = col_date(format = "%m/%d/%Y")))
skim_without_charts(dailyActivity)
```

Already loaded the dates in the correct format. Since apart from date and Id all columns are numeric, it makes sense to use skim_without_charts which gives a statistical summary of each variable. The function output is rendered as an html table, which is quite nice.

```{r summary}
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
#plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
