---
title: "Data Analysis"
author: "Jeronimo Miranda"
date: '2023-05-18'
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
```

```{r message=FALSE, include=FALSE, results='hide'}
knitr::knit('1_Data_Cleaning_and_manipulation.rmd', tangle=TRUE)
source('1_Data_Cleaning_and_manipulation.R')
```

```{r}
library(changepoint)
library(ggplot2)
library(dplyr)
library(tidyr)
library(knitr)
```


## Preliminary

For the analysis, we use the clean data objects from the (Cleaning and processing step)[https://github.com/jerolon/FitBit_Tracker_Data_Analytics_Practice/blob/main/1_Data_Cleaning_and_manipulation.md]:

- dailyActivity, hourly2dailyActivity
- heartrate_minutes
- hourlyActivity, minute2hour
- minuteActivity, minuteAct_BPM
- sleepDay
- weightLogInfo

## Types of activities

I would like to infer the kind of activities that the users are doing during the study. First, we want to check the daily data for a clue. Because we have "Distance" and "Time" in the form of minutes, it is irresistible to calculate velocities. We graph the velocity for each intensity class.

```{r velocities, echo=FALSE, warning=FALSE}
dailyActivity %>% mutate(HighVelocity = 60 * VeryActiveDistance/VeryActiveMinutes,  ModerateVelocity = 60 * ModeratelyActiveDistance/FairlyActiveMinutes, LightVelocity = 60 *  LightActiveDistance/LightlyActiveMinutes, SedentaryVelocity = 60 * SedentaryActiveDistance/SedentaryMinutes) %>% 
    #it has to be transformed to long format
    pivot_longer(names_to = "Intensity_Class", values_to = "Velocity", cols = ends_with("Velocity")) %>% 
    #Boxplot
    ggplot(aes(x = Intensity_Class, y = Velocity)) + geom_jitter(alpha = 0.2) + geom_boxplot(outlier.shape = NA) + theme_bw() + labs(title = "Velocity vs intensity of activity", subtitle = "Intensity classes have a distinct velocity", y = "Velocity km/h", x = "Intensity class")
```

This is at the same time a sanity check (no one is travelling at 100 km/h, no sedentary velocity is significantly above 0), and an overview of the kind of activities people are doing. A few jog (> 6.5 km/h). Most walk (or swim?). 

Although the velocities are very well defined for each category, there is quite the overlap. There are even some days with high or moderate intensity but zero velocity. 
These are probably people on a stationary bike, or lifting weights. Lower than average velocities could be people hiking, swimming or going upstairs: intense activities that do not cover much distance. This is hard to tell, however, because it is taking all the minutes in a day, whereas a workout can combine periods of different intensity activity.

We will use the data frame with minutes resolution to construct an "Activities" data frame. Here, I want to aggregate the minutes in a user day that have the same excercise intensity. To find the times when intensity mean changes, we use the changepoint algorigthm mentioned [here](https://rpubs.com/richkt/269908). We just pass the whole column, since it is already arranged by Id and DateTime.

```{r, echo=TRUE}
minuteAct_BPM <- arrange(minuteAct_BPM,Id,ActivityMinute )
#find the positions where the mean intensity changes
changepoint_position <- cpt.mean(minuteAct_BPM$Intensity,penalty='Manual',pen.value=8,method='PELT') %>% cpts()

minuteAct_BPM$IntensityChange <- rep(FALSE, nrow(minuteAct_BPM))
minuteAct_BPM$IntensityChange[changepoint_position + 1] <- TRUE

```

Now we have a column `IntensityChange` with value `TRUE` whenever we detected a change in activity levels. Now we calculate another column which gives us information on the time gap between records. We use the combination of these two columns to cluster Activities by clustering groups of adjacent records that have the same mean intensity and do not have temporal jumps bigger than 10 minutes (because the user took off the tracker or it is just a different day).

```{r, echo=TRUE}
activities <- minuteAct_BPM %>% 
  #Will be true when activity changes
  mutate(#Timegap is true whenever there is a gap bigger than 10 minutes
         timegap = minutes(10) < abs(ActivityMinute - lag(ActivityMinute, def = first(ActivityMinute))),
         #ActivityChange will only be true when either the activity changes or there is a big time jump
         ActivityChange = IntensityChange | timegap,
         activityId = cumsum(ActivityChange)
         ) %>% select(-timegap, -ActivityChange, -IntensityChange)

n_distinct(activities$activityId)

```

The data frame `activities` contains grouped spans of time that have the same mean intensity. To illustrate what I did, I plot the intensity for every minute in a day from the user 8877689391. As you can see, the intensity fluctuates normally. The red line is the mean intensity for each activity (This day is divided into 7 activities). Around 12 pm there is a long and very intense activity! Later, right after 15 there is an activity comprised of minutes with 1,2, and 3 intensities. These fluctuations are grouped together and the mean intensity of this activity is 2.

```{r, echo=FALSE}
activities %>% group_by(activityId) %>% mutate(dia = date(ActivityMinute), meanIntensity = mean(Intensity)) %>% filter((Id == "8877689391" & dia == "2016-04-10")) %>% ggplot() + geom_line(aes(x = ActivityMinute, y = Intensity)) + geom_line(aes(x = ActivityMinute, y = meanIntensity), col = "red", size = 1) + theme_bw() + labs(title = "A day in the life", subtitle = "Minutes grouped into activities by their intensity", x = "Time of day")

```


Now we can explore which properties the different activities have. In the below graph we plot the mean cadence (how many steps per minute) and the mean heart rate. Each dot is a single activity. The heart rate does not really correlate with the cadence (there are many ways to get your heart up without walking). The plot is faceted into different activity categories based on mean intensity.

- Sedentary activities (Intensity mean 0) are defined by less than 120 beats per minute and less than 25 steps per minute (This is a step every two seconds, so a bit active but not a continuous walk).
- Moderate intensity (1) activities are very clustered between 0-50 steps per minute and 60-120 bpm, again, if you can imagine taking less than 1 step every second, this is not a walk. This are mainly standing activities with sporadic walking, maybe cooking, working, or chores.
- Fairly intense activities (2) are defined by 80-120 BPM and less than 120 Steps per minute. I like how this category is so well defined by a BPM band, although it spans a wide spectrum in cadence.
- Last, the high intensity activities (3) are defined by either a > 100 BPM or >80 BPM combined with a cadence greater than 100 steps per minute. This suggests that we could subdivide this category further into brisk walking, and intense stationary exercise this could be anything from swimming, weightlifting or cycling (which the accelerometer does not count as steps). There is also a very natural division at cadence > 120 which we will call *jogging*.

```{r echo=FALSE}
activities %>% group_by(activityId) %>% summarise(
  #Just divide the max time for the min time and normalize to minutes
  duration = (max(ActivityMinute) - min(ActivityMinute))/dminutes(1),
  #Activities with only one minute
  duration = if_else(duration == 0,1,duration),
  #cadence is the number of steps per minute
  cadence = sum(Steps)/duration,  hrate = mean(BPM),
  Int = mean(Intensity),
  #Using "round" because still want to keep it as factor
  Intensity_class = round(Int)) %>% ggplot() +
  geom_point(aes(cadence, hrate, col = Int)) + theme_bw() + facet_wrap(~Intensity_class) + labs (title = "Overlap of cadence and heart rate in different activity classes", y = "Mean heart rate (BPM)", x = "Cadence (Steps/Minute)") + scale_fill_brewer(name = "Mean Intensity", palette = "Dark2")
```

We will now explore intensities 0, 2, and 3 further and subdivide them. Light intensity activities (1) are very well clustered. I am tempted to name them as "Chores" or "Daily activities".

#### Sedentary activities
More than half of activities are classified as sedentary. This could be sleep activities or just activities sitting down. We classify as "Deep Rest" whenever someone has a heart rate below 59 or average activity intensity below 0.15. Anything else is classified as "Rest". Below we plot all the activities properties against each other and we see how these two criteria are good at discriminating between groups of points. Red are "Deep rest" activities and blue are "Rest" activities. I specially like how the biggest peak for Deep rest is at ~600 minutes, which could be the average sleep time. 

```{r sedentary activities,echo=FALSE}
library(GGally)
activities %>% group_by(activityId) %>% summarise(
    #Just divide the max time for the min time and normalize to minutes
    duration = (max(ActivityMinute) - min(ActivityMinute))/dminutes(1),
    #Activities with only one minute, should probably be deleted
    duration = if_else(duration == 0,1,duration),
    #cadence is the number of steps per minute
    cadence = sum(Steps)/duration,  hrate = mean(BPM),
    Int = mean(Intensity), calories = sum(Calories),
    #Using "round" because still want to keep it as factor
    Intensity_class = round(Int)) %>%
    filter(Intensity_class == 0) %>% 
  #classify into deep rest and rest
  mutate(activity_name = if_else((hrate > 59 | Int > 0.15), "Rest", "Deep Rest")) %>% 
    ggpairs(columns = 2:6, aes(color = as.factor(activity_name), alpha = 0.5)) + theme_bw() + labs(title = "Pairwise comparisons of properties of sedentary activities")
```

Looking only at the hour of the day at which the different activities start, we have confidence that we are making a good subdivision. Most "Deep rest" starts at 8PM, while most "Rest" activities start between 6 AM and 8 PM.

```{r,echo=FALSE}
activities %>% group_by(activityId) %>% summarise(start = hour(min(ActivityMinute)),
    hrate = mean(BPM),
    Int = mean(Intensity),
    #Using "round" because still want to keep it as factor
    Intensity_class = round(Int)) %>%
    filter(Intensity_class == 0) %>% mutate(activity_name = if_else((hrate > 59 | Int > 0.15), "Rest", "Deep Rest")) %>% ggplot() + geom_histogram(aes(start), bins = 24) + facet_wrap(~activity_name, scales = "free_y") + theme_bw() + labs(title = "Hour that sedentary activities start")
```

#### Fairly intense activities
The next intensity class is well clustered and it is hard to subdivide. Cadence and heart-rate are pretty flat. Duration is tipically 10 minutes, seldom more than an hour. There is, however, a funny pattern when plotting mean intensity vs duration: when intensity is lower than 2, duration tends to rise with intensity, I call this cluster "Stroll", people are enjoying themselves. Higher than 2, intensity drops with duration, because the peak density cadence of this cluster is higher than 90, I call this cluster "Brisk walk". 

```{r,echo=FALSE}
activities %>% group_by(activityId) %>% summarise(
    #Just divide the max time for the min time and normalize to minutes
    duration = (max(ActivityMinute) - min(ActivityMinute))/dminutes(1),
    #Activities with only one minute, should probably be deleted
    duration = if_else(duration == 0,1,duration),
    #cadence is the number of steps per minute
    cadence = sum(Steps)/duration,  hrate = mean(BPM),
    Int = mean(Intensity), calories = sum(Calories),
    #Using "round" because still want to keep it as factor
    Intensity_class = round(Int)) %>%
    filter(Intensity_class == 2) %>% 
  mutate(activity_name = if_else((Int > 2), "Medium walk", "Stroll")) %>% 
    ggpairs(columns = 2:6, aes(color = as.factor(activity_name),alpha = 0.5), upper = list(continuous  = "points")) + theme_bw()
```

While strolls happen throughout the day, medium walks happen mostly in the afternoon.

```{r,echo=FALSE}
activities %>% group_by(activityId) %>% summarise(start = hour(min(ActivityMinute)),
    Int = mean(Intensity),
    #Using "round" because still want to keep it as factor
    Intensity_class = round(Int)) %>%
    filter(Intensity_class == 2) %>% 
    mutate(activity_name = if_else((Int > 2),  "Medium walk", "Stroll")) %>% ggplot() + geom_histogram(aes(start), bins = 24) + facet_wrap(~activity_name, scales = "free_y") + theme_bw()
    
```

#### High intensity activity

For the highest intensity class, cadence offers a very clear way to subdivide activities. A cadence below 60 is classified as a 
"workout" which is any intense exercise that does not involve much walking. A cadence between 60-120 is a brisk walk, a walk that is challenging, even if it is not so fast. It could include people walking upstairs or hiking. The last category is "Jogging", where people are going more than 120 steps per minute, these tend to be the most intense, but are not so numerous. 


```{r,echo=FALSE}
activities %>% group_by(activityId) %>% summarise(
    #Just divide the max time for the min time and normalize to minutes
    duration = (max(ActivityMinute) - min(ActivityMinute))/dminutes(1),
    #Activities with only one minute, should probably be deleted
    duration = if_else(duration == 0,1,duration),
    #cadence is the number of steps per minute
    cadence = sum(Steps)/duration,  hrate = mean(BPM),
    Int = mean(Intensity), calories = sum(Calories),
    #Using "round" because still want to keep it as factor
    Intensity_class = round(Int)) %>%
    filter(Intensity_class == 3) %>% 
    #classify into deep rest and rest
    mutate(activity_name = case_when(cadence < 60 ~ "Workout", between(cadence,60,120) ~ "Fast walk", cadence > 120 ~ "Jogging")) %>% 
    ggpairs(columns = 2:6, aes(color = as.factor(activity_name), alpha = 0.5), upper = list(continuous  = "points")) + theme_bw()
```

```{r include=FALSE,echo=FALSE}
blank_theme <- theme_minimal()+
  theme(
  axis.title.x = element_blank(),
  axis.title.y = element_blank(),
  panel.border = element_blank(),
  panel.grid=element_blank(),
  axis.ticks = element_blank(),
  plot.title=element_text(size=14, face="bold")
  )
```

The following code classifies the different activities according to the criteria outlined above. It could be fun to use machine learning algorithms to the activity data but this is beyond the business task right now and I don't think it would result in a significantly better classification.

```{r echo=TRUE}
activity_summary <- activities %>% group_by(activityId) %>% summarise(Id = unique(Id),
    #Just divide the max time for the min time and normalize to minutes
    duration = (max(ActivityMinute) - min(ActivityMinute))/dminutes(1),
    #Activities with only one minute, should probably be deleted
    duration = if_else(duration == 0,1,duration),
    #cadence is the number of steps per minute
    cadence = sum(Steps)/duration,  hrate = mean(BPM),
    Int = mean(Intensity), calories = sum(Calories),
    #Using "round" because still want to keep it as factor
    Intensity_class = round(Int)) %>%
    #Separate by intensity class, then the conditions for each sub-class
  mutate(activity_name = case_when(
    Intensity_class == 3 ~ case_when(cadence < 60 ~ "Workout", between(cadence,60,120) ~ "Fast walk", cadence > 120 ~ "Jogging"),
    Intensity_class == 2 ~ if_else((Int > 2),  "Medium walk", "Stroll"),
    Intensity_class == 1 ~ "Normal activity",
    Intensity_class == 0 ~ if_else((hrate > 59 | Int > 0.15), "Rest", "Deep Rest"),
    TRUE ~ "Unclassified activity")) 
```

Most of the time, people spend either resting or going normally about their day. Not in physical activity. Therefore, we filter out the sedentary activities and normal life to get only physical activities, however mild. 


```{r echo=FALSE}
library("gridExtra")
pie <- activity_summary %>% group_by(activity_name) %>% summarise(number = n()) %>% filter(!stringr::str_ends(activity_name, "Rest"), activity_name != "Normal activity") %>%  ggplot(aes(x="", y = number, fill = activity_name)) + geom_bar(width = 1, stat = "identity") + coord_polar("y", start = 0) + scale_fill_brewer(name = "Activity", palette = "Dark2") + theme_minimal() + blank_theme + theme(axis.text.x=element_blank()) + labs(title = "Tipes of activities of fitbit users", subtitle = "Fitbit user activity is dominated by walking")

bar <- activity_summary %>% group_by(Id,activity_name) %>% summarise(number = n()) %>% filter(!stringr::str_ends(activity_name, "Rest"), activity_name != "Normal activity") %>% mutate(perc = 100 * number/sum(number)) %>%  ggplot(aes(y=Id, x = perc, fill = activity_name)) + geom_bar(width = 0.7, stat = "identity") + scale_fill_brewer(palette = "Dark2") + theme_bw() + theme(axis.text.y = element_blank(), legend.position="none") + labs(subtitle = "Activity type by user", x = "Percentage", y = "User")

grid.arrange(pie, bar, ncol = 2)
```

A single user could be defined as a runner. Another user workouts in more than 50% of their activities. However, the overall pattern is that most activities done by people who use fitness trackers involve walking with different intensities.

#### Activity Logging

We take a small detour to explore another trend in fitbit usage. We examine the proportion of days in which users log an activity manually. Overwhelmingly, users track their activities automatically. The total distance that is manually logged is low, suggesting that this is not a popular feature. 

```{r echo=FALSE}
dailyActivity %>% mutate(Logged = if_else(LoggedActivitiesDistance > 0, "Logged Activities", "No Logged Activities")) %>% group_by(Logged) %>% summarise(number = n()) %>%  ggplot(aes(x="", y = number, fill = Logged)) + geom_bar(width = 1, stat = "identity") + coord_polar("y", start = 0) + scale_fill_brewer(name = "Activity", palette = "Dark2") + theme_minimal() + blank_theme + theme(axis.text.x=element_blank()) + labs(title = "Minimal use of Logged activities feature", subtitle = "Days with any Logged activities distance")
```

This does not mean that you should eliminate the logged activities feature. Although only 6 users logged any activity distance at all, for these 6, the distance tracked by manual logging represents a substantial percentage of the total distance they track. Moreover, even these users do not log activities everyday, but they log a greater total distance on the days that they manually log their activities.

```{r, echo=FALSE}
a <- dailyActivity %>% mutate(manual = 100 * LoggedActivitiesDistance / (LoggedActivitiesDistance + TotalDistance)) %>% group_by(Id) %>% filter(any(manual > 0)) %>% ggplot(aes(y = Id, x = manual)) + geom_boxplot(orientation = "y") + theme_bw() + labs(title = "A small number of of users Manually Logs their activities", subtitle = "Percent of Logged Activities Distance of the users that use the feature", x = "Logged distance as % of day's total")

#The users that have any days with logged activities
manualLoggers <- dailyActivity %>% filter(any(LoggedActivitiesDistance > 0)) %>% pull(Id) %>% unique()

b <- dailyActivity %>% filter(Id %in% manualLoggers) %>% ggplot(aes(x = LoggedActivitiesDistance, y = TotalDistance)) + geom_point(aes(col = Id)) + geom_smooth(method = "glm", col = "red") + theme_bw() + labs(title = "Logging can help being more active", subtitle = "Correlation between daily logged distance and total distance", x = "Daily distance logged", y = "Total daily distance") + theme(legend.position="none")

grid.arrange(a, b, nrow = 2)
```


### Sedentarism

What's the relationship between steps taken in a day and sedentary minutes? How could this help inform the customer segments that we can market to? E.g. position this more as a way to get started in walking more? Or to measure steps that you're already taking?

```{r echo=FALSE}
ggplot(data=dailyActivity, aes(x=TotalSteps, y=SedentaryMinutes)) + geom_point() + theme_bw()
```

This is very interesting, it suggests that we can segment users by those that have more than 1000 sedentary minutes a day and those that have less, lets do that. We will put the users who have more than 1000 Sedentary minutes more than half their dates on a group called "Sedentary".

```{r}
dailyActivity <- dailyActivity %>% group_by(Id) %>% mutate(User_Segment = if_else(median(SedentaryMinutes) > 1000, "Sedentary", "Active"))
```

Now, sedentary users might want to minimize their total sedentary minutes, at least to less than 1000, which would put them on the "Active" group. We can check which of the variables minimizes the sedentarism, in order to better help this group reach their goals. We will plot the sedentary minutes against other variables and calculate negative correlations. When we take a look at the left most column and the top row, we see that, although total steps and sedentary minutes are negatively correlated, the strongest negative correlation is LightlyActive Minutes!

```{r echo=FALSE}
dailyActivity %>% relocate(SedentaryMinutes, .after = ActivityDate) %>% ggpairs(columns = c(3,4,8:10,12:14),aes(color = as.factor(User_Segment)))
```

This is expected -- because a day only has 1440 minutes, if you spend at least 440 minutes being a little active, then you cannot spend more than 1000 minutes inactive. Still, it suggests that for people that are struggling to be more active, an emphasis on time doing some activity, any activity, is likely to help them more achieve their goals, even on days when they cannot reach the specified number of steps. 
When you look at the bottom row of the graph below, it is clear that "Sedentary" users are walking more than 5000 or even 10000 steps sometimes. Yet, despite this, they are still very sedentary most days. This could suggest that an emphasis on doing 10,000 steps a day might discourage people who do not have the time every day, or that once they achieve this, they are too tired to do much else.

```{r echo=FALSE, fig.height=5, fig.width=7}
a <- ggplot(data=dailyActivity, aes(x=LightlyActiveMinutes, y=SedentaryMinutes)) + geom_point(aes(col = User_Segment)) + geom_smooth(method = "glm", aes(col = User_Segment)) + theme_bw() + labs(title = "Being even slightly active is a good start", subtitle = " Relationship between light activity and sedentarism") + facet_wrap(~ User_Segment) + geom_hline(aes(yintercept=1000), linetype = "dashed", col = "red", size = 1) +  theme(legend.position="none")

b <- ggplot(data=dailyActivity, aes(x=TotalSteps, y=SedentaryMinutes)) + geom_point(aes(col = User_Segment)) + geom_smooth(method = "glm", aes(col = User_Segment)) + theme_bw() + labs(title = " Achieving more than a 1000 steps is no guarantee", subtitle = "Relationship between total steps and sedentary minutes") + facet_wrap(~ User_Segment) + geom_hline(aes(yintercept=1000), linetype = "dashed", col = "red", size = 1) + theme(legend.position="none")

grid.arrange(a, b, nrow = 2)
```

The great news is that, for users who are looking to be more active and burning more calories, the relationship between steps and calories is almost strictly linear. The benefits do not stop: 5,000 steps is better than 0 steps, 10,000 steps is even better better, and the calories burned keep increasing with the number of steps. You can still set a target number of steps, but you should know that whatever activity you do is valuable to stay healthy.


```{r, echo=FALSE}
ggplot(data=dailyActivity, aes(x=TotalSteps, y=Calories)) + geom_point() + geom_smooth(method = "glm", col ="red") + theme_bw() + labs(title = "More steps = More calories", subtitle = "A linear relationship between total steps and burned calories") + facet_wrap(~User_Segment)
```

### Resting heart rate

One interestign parameter that we can estimate is *resting heart rate*. Although it can vary by age, in general, having a lower heart rate implies more efficient heart function and better cardiovascular fitness. We estimate it by filtering all the minutes with activity intensity 0 ("rest") in a day and taking the mean. We then join it with the dailyActivity data frame to compare it with the number of steps. We do inner join to keep only records that are in both data frames. Afterwards, we summarise again to get a single number for every Id.

```{r echo=TRUE}
day_rest_HR <- minuteAct_BPM %>% mutate(ActivityDate = date(ActivityMinute)) %>%
  group_by(Id, ActivityDate) %>% filter(Intensity == 0) %>% 
  summarise(Id = unique(Id), ActivityDate = unique(ActivityDate), resting_heart_rate = mean(BPM))

rhr_by_user <- inner_join(dailyActivity, day_rest_HR, by = c("Id", "ActivityDate")) %>% group_by(Id) %>%
  summarise(Id = unique(Id),
            across(c(resting_heart_rate, TotalSteps, VeryActiveDistance:LightActiveDistance, ends_with("Minutes"), Calories),
                   #just means that we apply the function mean to all the above columns
                   mean)
            
            )


```


Users that walk more tend to have a lower resting heart rate, but wait, the three users with the lowest resting heart rate are not the ones that take the most steps (gray rectangle). These same three users are also among the ones with the least sedentary minutes (bottom left). The strongest anti-correlation is with Very active minutes, suggesting that the intensity of activities is more important than the number of steps. Unfortunately, lightly active minutes has almost no bearing on cardiovascular health. Overall, no activity explains everything because it is a small sample and a noisy measurement. However, the correlations are clear: move or walk intensely and try to not spend much time being sedentary for better health.

```{r, echo=FALSE}
g1 <- rhr_by_user %>% ggplot(aes(x = TotalSteps, y = resting_heart_rate)) + geom_point(aes(col=Id)) + geom_smooth(col = "red", method = "glm", se = FALSE) + annotate("text", x = 15000, y = 80, label = paste0("italic(R)^2 == ", cor(rhr_by_user$TotalSteps, rhr_by_user$resting_heart_rate)), parse = TRUE) + theme_bw()  + theme(legend.position="none") + labs(title = "Correlations of resting heart rate with activity by user", y = "Mean Resting Heart Rate", x = "Mean of daily Total Steps") + annotate(geom = "rect", xmin = 8000, xmax = 12000, ymin = 55, ymax = 65, alpha = 0.2)

g2 <- rhr_by_user %>% ggplot(aes(x = VeryActiveMinutes, y = resting_heart_rate)) + geom_point(aes(col=Id)) + geom_smooth(col = "red", method = "glm", se = FALSE) + annotate("text", x = 50, y = 80, label = paste0("italic(R)^2 == ", cor(rhr_by_user$VeryActiveMinutes, rhr_by_user$resting_heart_rate)), parse = TRUE) + theme_bw()  + theme(legend.position="none") + labs(y = "Mean Resting Heart Rate", x = "Mean of daily Very Active Minutes")

g4 <- rhr_by_user %>% ggplot(aes(x = LightlyActiveMinutes, y = resting_heart_rate)) + geom_point(aes(col=Id)) + geom_smooth(col = "red", method = "glm", se = FALSE) + annotate("text", x = 250, y = 80, label = paste0("italic(R)^2 == ", cor(rhr_by_user$LightlyActiveMinutes, rhr_by_user$resting_heart_rate)), parse = TRUE) + theme_bw()  + theme(legend.position="none") + labs(y = "Mean Resting Heart Rate", x = "Mean of daily lightly Active Minutes")

g3 <- rhr_by_user %>% ggplot(aes(x = SedentaryMinutes, y = resting_heart_rate)) + geom_point(aes(col=Id)) + geom_smooth(col = "red", method = "glm", se = FALSE) + annotate("text", x = 1200, y = 65, label = paste0("italic(R)^2 == ", cor(rhr_by_user$SedentaryMinutes, rhr_by_user$resting_heart_rate)), parse = TRUE) + theme_bw()  + theme(legend.position="none") + labs(y = "Mean Resting Heart Rate", x = "Mean number of sedentary Active Minutes") + annotate("rect", xmin = 650, xmax = 810, ymin = 55, ymax = 65, alpha = 0.2)

grid.arrange(g1,g2,g3,g4, nrow = 2)
```


## Sleep and activity

We have sleep records for a subset of users. What's the relationship between minutes asleep and time in bed? You might expect it to be almost completely linear - and it is. But there are examples of people that are spending a typical amount of time in bed (6 to 10 hours), but are getting less sleep than they think, and usually less than 7.5 hours, which is recommended. 

```{r}
ggplot(data=sleepDay, aes(x=TotalMinutesAsleep/60, y=TotalTimeInBed/60)) + geom_point() + theme_bw() + labs(x = "Hours asleep", y = "Time in bed") + annotate("rect", xmin = 3.4, xmax = 7.5, ymin = 5, ymax = 11, alpha = 0.2) + scale_x_continuous(breaks = seq(0, 16, by = 2)) + scale_y_continuous(breaks = seq(0, 16, by = 2)) + geom_abline(col = "red", intercept = 0, slope = 1)
```


Do participants who sleep more also take more steps or fewer steps per day? Is there a relationship at all? How could these answers help inform the marketing strategy of how you position this new product?

Amazingly, there is only a week correlation between total steps and minutes asleep!
And this correlation is negative, that is the more steps you take, the fewer minutes you sleep. What is more surprising is the anti-correlation with the number of sedentary minutes! I would have expected completely the opposite, since sleep counts into sedentary minutes (you are not very active when you are asleep), the more you sleep, the more sedentary minutes you would spend. 

It is hard to infer causation just from this correlation: on one hand, having too little sleep can make you less active because you are tired; on the other hand, not having enough activity can result in not being able to fall asleep at night. In any case, the graphs further our conclusion that the number of sedentary minutes is an important metric to monitor your general health, including sleep. 

```{r echo=FALSE}
#Inner join to keep only records that are present in both
sleep_joined <- inner_join(sleepDay, dailyActivity, by = join_by(Id, SleepDay == ActivityDate))

sleep1 <- sleep_joined %>% ggplot(aes(x = TotalSteps, y = TotalMinutesAsleep)) + geom_point() + geom_smooth(col = "red", method = "glm") + annotate("text", x = 15000, y = 700, label = paste0("italic(R)^2 == ", cor(sleep_joined$TotalSteps, sleep_joined$TotalMinutesAsleep)), parse = TRUE) + theme_bw()

sleep2 <- sleep_joined %>% ggplot(aes(x = SedentaryMinutes, y = TotalMinutesAsleep)) + geom_point() + geom_smooth(col = "red", method = "glm") + annotate("text", x = 1000, y = 700, label = paste0("italic(R)^2 == ", cor(sleep_joined$SedentaryMinutes, sleep_joined$TotalMinutesAsleep)), parse = TRUE) + theme_bw()

grid.arrange(sleep1, sleep2, nrow = 2)
```

## Conclusions

In summary, although some specialized users use the tracker as a monitor for running, the main trend of most users is to wear the device as they go through their daily activities. Although users incorporate running into their workouts, they do so very infrequently. Instead, most people activity consists on walking at a slow to moderate pace, as well as workouts that do not involve step count. The emphasis on step count is justified, because most activities involve walking in different intensities. However, many healthy activities cannot be detected by step count: yoga, lifting weights, cycling, but are still detected as active minutes. Moreover, other parameters are predictive of good health. Therefore, most users would benefit from the tracker as a general monitor  of activity levels, calories, sleep, heart rate and more. The device can be marketed as a general companion that gives you feedback on your health goals, that are as diverse as users.
